---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r setup}
options(scipen=999) #to remove scientific notation
setwd("~/Dropbox/RTutorials/Coursera5-ReproducibleResearch/RepData_PeerAssessment1")

#if the activity file hasn't been unzipped, unzip it
if (!("activity.csv" %in% list.files(".")))  unzip("activity.zip")

data <- read.csv("activity.csv", header = TRUE,
                 stringsAsFactors = FALSE)

#make date variable of class date
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?

```{r part_1}
library(plyr)
library(ggplot2)

#calculate the total number of steps for each date, excluding NA values
steps_day <- ddply(data, .(date), summarise,
                   totalSteps = sum(steps, na.rm = TRUE))

#create histogram using ggplot & geom_histogram
ggplot(steps_day, aes(x=totalSteps)) + 
    geom_histogram(binwidth=2000, colour="black", fill="white") +
    xlab("Total Steps Per Day") + ylab("Count") +
    ggtitle("Histogram of Total Steps Per Day - Missing Values Removed")+
    theme_bw()

#calculate mean and median number of steps across days
meanSteps <- mean(steps_day$totalSteps, na.rm = TRUE)
meanSteps

medianSteps <- median(steps_day$totalSteps, na.rm = TRUE)
medianSteps
```

The mean number of steps taken per day is `r round(meanSteps,0)`, excluding the missing values, while the median number of steps taken per day is `r medianSteps`.  

## What is the average daily activity pattern?

```{r part_2}
#calculate the average number of steps for each interval, across dates
#excluding NA values
steps_int <- ddply(data, .(interval), summarise,
                   meanSteps = mean(steps, na.rm = TRUE),
                   sdSteps = sd(steps, na.rm = TRUE))

#plot the average daily activity pattern using ggplot & geom_line
ggplot(steps_int, aes(y=meanSteps, x=interval, group = 1)) + 
    geom_line(color = "blue", alpha = 0.7) +
    xlab("Interval") + ylab("Mean Number of Steps") +
    ggtitle("Mean Number of Steps Per Interval")+
    theme_bw()

#calculate the max value, use which.max to get the index for the max
#and use the index to get the interval with the max number of steps
maxValue <- max(steps_int$meanSteps)
maxIndex <- which.max(steps_int$meanSteps)
maxInterval <- steps_int[maxIndex,1]

maxValue
maxInterval
```

The `r maxInterval`th interval is the five-minute interval containing the maximum number of steps (on average across all days in the data set), with `r round(maxValue,0)` steps.  

## Imputing missing values

Here I will impute missing values based on the average for the given interval across the other days that did not have missing values (calculated in the previous step).  

```{r part_3}
#get index of rows containing NA values for steps 
ind <- which(is.na(data), arr.ind = TRUE)
ind_rows <- ind[,1]

#create a vector int_means with the averages for each interval across days
#repeated 61 times to line up with the intervals for each day
int_means <- rep(steps_int$meanSteps,61)

#copy data into data_filled and then
#replace NA values (as indicated by ind_rows) with means from int_means
data_filled <- data
data_filled$steps[ind_rows] <- int_means[ind_rows]

#calculate the total number of steps for each date
steps_day_filled <- ddply(data_filled, .(date), summarise,
                   totalSteps = sum(steps))

#create histogram using ggplot & geom_histogram
ggplot(steps_day_filled, aes(x=totalSteps)) + 
    geom_histogram(binwidth=2000, colour="black", fill="white") +
    xlab("Total Steps Per Day") + ylab("Count") +
    ggtitle("Histogram of Total Steps Per Day - Missing Values Imputed")+
    theme_bw()

#calculate the mean and median number of steps across days
meanSteps_filled <- mean(steps_day_filled$totalSteps)
meanSteps_filled

medianSteps_filled <- median(steps_day_filled$totalSteps)
medianSteps_filled
```

With the imputed values, the mean and median are equivalent: mean = `r round(meanSteps_filled,0)` and median = `r round(medianSteps_filled,0)`. By imputing the missing values using the mean number of steps for the given interval across the other days, the mean and median number of steps taken per day has increased from the estimates calculated when we excluded missing values (estimates excluding missing values: mean = `r round(meanSteps,0)` and median = `r medianSteps`).

## Are there differences in activity patterns between weekdays and weekends?

```{r part_4, message=FALSE}
library(dplyr)
#create a variable with the actual weekday label to confirm 
#that weekend/weekday classification is done correctly
data_filled$date_type <- weekdays(data_filled$date)

#create a variable to label weekend/weekday
data_filled <- mutate(data_filled, 
                       Weekday_Type = (ifelse(weekdays(date) %in% c("Saturday", "Sunday"), 
                                              "Weekend", "Weekday")))

#re-order the weekday type variable to later make Weekend appear above on the graph
data_filled$Weekday_Type <- factor(data_filled$Weekday_Type, levels = c("Weekend", "Weekday"))

#calculate the mean number of steps as a function of interval and weekday type
steps_int_weekend <- ddply(data_filled, .(interval,Weekday_Type), 
                           summarise,
                           meanSteps = mean(steps))

#plot the average daily activity pattern using ggplot & geom_line
#use facet_grid to plot weekday and weekend separately
ggplot(steps_int_weekend, aes(y=meanSteps, x=interval, group = 1)) + 
    geom_line(color = "blue", alpha = 0.7) +
    facet_grid(Weekday_Type~.) +
    xlab("Interval") + ylab("Mean Number of Steps") +
    ggtitle("Mean Number of Steps Per Interval for Weekends and Weekdays")+
    theme_bw()

#Examine differences in max, min, and mean values between weekdays and weekends
weekend_step_data <- group_by(steps_int_weekend, Weekday_Type)
weekend_summary <- summarise(weekend_step_data, Max = max(meanSteps), 
          Min = min(meanSteps), 
          Mean = mean(meanSteps))
weekend_summary

```

In comparing activity patterns between weekdays and weekends, an interesting pattern emerges. Although the maximum number of steps by interval is greater on weekdays (`r round(weekend_summary[weekend_summary$Weekday_Type == "Weekday",2],0)`) than on weekends (`r round(weekend_summary[weekend_summary$Weekday_Type == "Weekend",2],0)`), the average number of steps across intervals is greater on weekends (`r round(weekend_summary[weekend_summary$Weekday_Type == "Weekend",4],0)`) than on weekdays (`r round(weekend_summary[weekend_summary$Weekday_Type == "Weekday",4],0)`).
